### 2018-09-26 paper/paper.tex
Revisi 3

\documentclass[conference]{IEEEtran}
\IEEEoverridecommandlockouts
% The preceding line is only needed to identify funding in the first footnote. If that is unneeded, please comment it out.
\usepackage{cite}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{algorithmic}
\usepackage{graphicx}
\usepackage{textcomp}
\usepackage{xcolor}
\def\BibTeX{{\rm B\kern-.05em{\sc i\kern-.025em b}\kern-.08em
    T\kern-.1667em\lower.7ex\hbox{E}\kern-.125emX}}
\begin{document}

@@

\begin{abstract}
    Saat ini, banyak transaksi penting terjadi dunia maya. Keamanan informasi menjadi jaminan yang sangat penting agar data-data penting tidak bocor dan disalahgunakan. Sistem deteksi dan pencegahan intrusi telah dikembangkan. Namun, kecepatan deteksi intrusi belum secepat peningkatan kecepatan jaringan.
    Solusi yang dapat digunakan yaitu penggunaan \emph{multithread} untuk pencocokan yang banyak secara paralel. Solusi \emph{multithreading} telah dikembangkan menggunakan CPU. Namun, CPU memiliki \emph{core} yang sangat terbatas. Alternatif lainnya adalah dengan menggunakan GPU. GPU memiliki kemampuan untuk membangkitkan banyak \emph{thread} sekaligus. GPU sangat cocok untuk melakukan komputasi sederhana dalam jumlah besar.     
    Salah satu komponen penting dalam sistem deteksi intrusi yaitu pencocokan string. Beban pencocokan string ini seringkali menjadi \emph{bottleneck} dalam analisis paket jaringan. Tugas akhir ini akan mencoba melakukan eksperimen untuk melakukan implementasi pencocokan string pada sistem deteksi intrusi Snort menggunakan GPU.    
    Untuk dapat mempercepat proses analisis menggunakan GPU, translasi solusi yang ada tidaklah cukup. Operasi dalam GPU seringkali adalah operasi yang I/O \emph{bound} dan \emph{memory bound}. Diperlukan beberapa pengubahan seperti alokasi \emph{thread} yang berbeda, metode transfer memori \emph{host} dan \emph{device}, skema penampungan paket, dan pengubahan struktur data \emph{state machine}. Implementasi dapat menghasilkan \emph{speedup} yang signifikan hingga 3x lipat solusi \emph{multithreading} dengan CPU.
\end{abstract}

\begin{IEEEkeywords}
    \textit{pattern matching}, deteksi intrusi, GPU, optimasi, paralel
\end{IEEEkeywords}

@@

\section{Ease of Use}

\subsection{Maintaining the Integrity of the Specifications}

The IEEEtran class file is used to format your paper and style the text. All margins, 
column widths, line spaces, and text fonts are prescribed; please do not 
alter them. You may note peculiarities. For example, the head margin
measures proportionately more than is customary. This measurement 
and others are deliberate, using specifications that anticipate your paper 
as one part of the entire proceedings, and not as an independent document. 
Please do not revise any of the current designations.

\section{Prepare Your Paper Before Styling}
Before you begin to format your paper, first write and save the content as a 
separate text file. Complete all content and organizational editing before 
formatting. Please note sections \ref{AA}--\ref{SCM} below for more information on 
proofreading, spelling and grammar.

Keep your text and graphic files separate until after the text has been 
formatted and styled. Do not number text heads---{\LaTeX} will do that 
for you.

\subsection{Abbreviations and Acronyms}\label{AA}
Define abbreviations and acronyms the first time they are used in the text, 
even after they have been defined in the abstract. Abbreviations such as 
IEEE, SI, MKS, CGS, ac, dc, and rms do not have to be defined. Do not use 
abbreviations in the title or heads unless they are unavoidable.

\subsection{Units}
\begin{itemize}
\item Use either SI (MKS) or CGS as primary units. (SI units are encouraged.) English units may be used as secondary units (in parentheses). An exception would be the use of English units as identifiers in trade, such as ``3.5-inch disk drive''.
\item Avoid combining SI and CGS units, such as current in amperes and magnetic field in oersteds. This often leads to confusion because equations do not balance dimensionally. If you must use mixed units, clearly state the units for each quantity that you use in an equation.
\item Do not mix complete spellings and abbreviations of units: ``Wb/m\textsuperscript{2}'' or ``webers per square meter'', not ``webers/m\textsuperscript{2}''. Spell out units when they appear in text: ``. . . a few henries'', not ``. . . a few H''.
\item Use a zero before decimal points: ``0.25'', not ``.25''. Use ``cm\textsuperscript{3}'', not ``cc''.)
\end{itemize}

\subsection{Equations}
Number equations consecutively. To make your 
equations more compact, you may use the solidus (~/~), the exp function, or 
appropriate exponents. Italicize Roman symbols for quantities and variables, 
but not Greek symbols. Use a long dash rather than a hyphen for a minus 
sign. Punctuate equations with commas or periods when they are part of a 
sentence, as in:
\begin{equation}
a+b=\gamma\label{eq}
\end{equation}

Be sure that the 
symbols in your equation have been defined before or immediately following 
the equation. Use ``\eqref{eq}'', not ``Eq.~\eqref{eq}'' or ``equation \eqref{eq}'', except at 
the beginning of a sentence: ``Equation \eqref{eq} is . . .''

\subsection{\LaTeX-Specific Advice}

Please use ``soft'' (e.g., \verb|\eqref{Eq}|) cross references instead
of ``hard'' references (e.g., \verb|(1)|). That will make it possible
to combine sections, add equations, or change the order of figures or
citations without having to go through the file line by line.

Please don't use the \verb|{eqnarray}| equation environment. Use
\verb|{align}| or \verb|{IEEEeqnarray}| instead. The \verb|{eqnarray}|
environment leaves unsightly spaces around relation symbols.

Please note that the \verb|{subequations}| environment in {\LaTeX}
will increment the main equation counter even when there are no
equation numbers displayed. If you forget that, you might write an
article in which the equation numbers skip from (17) to (20), causing
the copy editors to wonder if you've discovered a new method of
counting.

{\BibTeX} does not work by magic. It doesn't get the bibliographic
data from thin air but from .bib files. If you use {\BibTeX} to produce a
bibliography you must send the .bib files.

{\LaTeX} can't read your mind. If you assign the same label to a
subsubsection and a table, you might find that Table I has been cross
referenced as Table IV-B3.

{\LaTeX} does not have precognitive abilities. If you put a
\verb|\label| command before the command that updates the counter it's
supposed to be using, the label will pick up the last counter to be
cross referenced instead. In particular, a \verb|\label| command
should not go before the caption of a figure or a table.

Do not use \verb|\nonumber| inside the \verb|{array}| environment. It
will not stop equation numbers inside \verb|{array}| (there won't be
any anyway) and it might stop a wanted equation number in the
surrounding equation.

\subsection{Some Common Mistakes}\label{SCM}
\begin{itemize}
\item The word ``data'' is plural, not singular.
\item The subscript for the permeability of vacuum $\mu_{0}$, and other common scientific constants, is zero with subscript formatting, not a lowercase letter ``o''.
\item In American English, commas, semicolons, periods, question and exclamation marks are located within quotation marks only when a complete thought or name is cited, such as a title or full quotation. When quotation marks are used, instead of a bold or italic typeface, to highlight a word or phrase, punctuation should appear outside of the quotation marks. A parenthetical phrase or statement at the end of a sentence is punctuated outside of the closing parenthesis (like this). (A parenthetical sentence is punctuated within the parentheses.)
\item A graph within a graph is an ``inset'', not an ``insert''. The word alternatively is preferred to the word ``alternately'' (unless you really mean something that alternates).
\item Do not use the word ``essentially'' to mean ``approximately'' or ``effectively''.
\item In your paper title, if the words ``that uses'' can accurately replace the word ``using'', capitalize the ``u''; if not, keep using lower-cased.
\item Be aware of the different meanings of the homophones ``affect'' and ``effect'', ``complement'' and ``compliment'', ``discreet'' and ``discrete'', ``principal'' and ``principle''.
\item Do not confuse ``imply'' and ``infer''.
\item The prefix ``non'' is not a word; it should be joined to the word it modifies, usually without a hyphen.
\item There is no period after the ``et'' in the Latin abbreviation ``et al.''.
\item The abbreviation ``i.e.'' means ``that is'', and the abbreviation ``e.g.'' means ``for example''.
\end{itemize}
An excellent style manual for science writers is \cite{b7}.

\subsection{Authors and Affiliations}
\textbf{The class file is designed for, but not limited to, six authors.} A 
minimum of one author is required for all conference articles. Author names 
should be listed starting from left to right and then moving down to the 
next line. This is the author sequence that will be used in future citations 
and by indexing services. Names should not be listed in columns nor group by 
affiliation. Please keep your affiliations as succinct as possible (for 
example, do not differentiate among departments of the same organization).

\subsection{Identify the Headings}
Headings, or heads, are organizational devices that guide the reader through 
your paper. There are two types: component heads and text heads.

Component heads identify the different components of your paper and are not 
topically subordinate to each other. Examples include Acknowledgments and 
References and, for these, the correct style to use is ``Heading 5''. Use 
``figure caption'' for your Figure captions, and ``table head'' for your 
table title. Run-in heads, such as ``Abstract'', will require you to apply a 
style (in this case, italic) in addition to the style provided by the drop 
down menu to differentiate the head from the text.

Text heads organize the topics on a relational, hierarchical basis. For 
example, the paper title is the primary text head because all subsequent 
material relates and elaborates on this one topic. If there are two or more 
sub-topics, the next level head (uppercase Roman numerals) should be used 
and, conversely, if there are not at least two sub-topics, then no subheads 
should be introduced.

\subsection{Figures and Tables}
\paragraph{Positioning Figures and Tables} Place figures and tables at the top and 
bottom of columns. Avoid placing them in the middle of columns. Large 
figures and tables may span across both columns. Figure captions should be 
below the figures; table heads should appear above the tables. Insert 
figures and tables after they are cited in the text. Use the abbreviation 
``Fig.~\ref{fig}'', even at the beginning of a sentence.

\begin{table}[htbp]
\caption{Table Type Styles}
\begin{center}
\begin{tabular}{|c|c|c|c|}
\hline
\textbf{Table}&\multicolumn{3}{|c|}{\textbf{Table Column Head}} \\
\cline{2-4} 
\textbf{Head} & \textbf{\textit{Table column subhead}}& \textbf{\textit{Subhead}}& \textbf{\textit{Subhead}} \\
\hline
copy& More table copy$^{\mathrm{a}}$& &  \\
\hline
\multicolumn{4}{l}{$^{\mathrm{a}}$Sample of a Table footnote.}
\end{tabular}
\label{tab1}
\end{center}
\end{table}

\begin{figure}[htbp]
\centerline{\includegraphics{fig1.png}}
\caption{Example of a figure caption.}
\label{fig}
\end{figure}

Figure Labels: Use 8 point Times New Roman for Figure labels. Use words 
rather than symbols or abbreviations when writing Figure axis labels to 
avoid confusing the reader. As an example, write the quantity 
``Magnetization'', or ``Magnetization, M'', not just ``M''. If including 
units in the label, present them within parentheses. Do not label axes only 
with units. In the example, write ``Magnetization (A/m)'' or ``Magnetization 
\{A[m(1)]\}'', not just ``A/m''. Do not label axes with a ratio of 
quantities and units. For example, write ``Temperature (K)'', not 
``Temperature/K''.

@@

\section*{References}

Please number citations consecutively within brackets \cite{b1}. The 
sentence punctuation follows the bracket \cite{b2}. Refer simply to the reference 
number, as in \cite{b3}---do not use ``Ref. \cite{b3}'' or ``reference \cite{b3}'' except at 
the beginning of a sentence: ``Reference \cite{b3} was the first $\ldots$''

Number footnotes separately in superscripts. Place the actual footnote at 
the bottom of the column in which it was cited. Do not put footnotes in the 
abstract or reference list. Use letters for table footnotes.

Unless there are six authors or more give all authors' names; do not use 
``et al.''. Papers that have not been published, even if they have been 
submitted for publication, should be cited as ``unpublished'' \cite{b4}. Papers 
that have been accepted for publication should be cited as ``in press'' \cite{b5}. 
Capitalize only the first word in a paper title, except for proper nouns and 
element symbols.

For papers published in translation journals, please give the English 
citation first, followed by the original foreign-language citation \cite{b6}.

\begin{thebibliography}{00}
\bibitem{b1} G. Eason, B. Noble, and I. N. Sneddon, ``On certain integrals of Lipschitz-Hankel type involving products of Bessel functions,'' Phil. Trans. Roy. Soc. London, vol. A247, pp. 529--551, April 1955.
\bibitem{b2} J. Clerk Maxwell, A Treatise on Electricity and Magnetism, 3rd ed., vol. 2. Oxford: Clarendon, 1892, pp.68--73.
\bibitem{b3} I. S. Jacobs and C. P. Bean, ``Fine particles, thin films and exchange anisotropy,'' in Magnetism, vol. III, G. T. Rado and H. Suhl, Eds. New York: Academic, 1963, pp. 271--350.
\bibitem{b4} K. Elissa, ``Title of paper if known,'' unpublished.
\bibitem{b5} R. Nicole, ``Title of paper with only first word capitalized,'' J. Name Stand. Abbrev., in press.
\bibitem{b6} Y. Yorozu, M. Hirano, K. Oka, and Y. Tagawa, ``Electron spectroscopy studies on magneto-optical media and plastic substrate interface,'' IEEE Transl. J. Magn. Japan, vol. 2, pp. 740--741, August 1987 [Digests 9th Annual Conf. Magnetics Japan, p. 301, 1982].
\bibitem{b7} M. Young, The Technical Writer's Handbook. Mill Valley, CA: University Science, 1989.
\end{thebibliography}
\vspace{12pt}
\color{red}
IEEE conference templates contain guidance text for composing and formatting conference papers. Please ensure that all template text is removed from your conference paper prior to submission to the conference. Failure to remove the template text from your paper may result in your paper not being published.

>>>>

\documentclass[conference]{IEEEtran}
\IEEEoverridecommandlockouts
% The preceding line is only needed to identify funding in the first footnote. If that is unneeded, please comment it out.
\usepackage[noadjust]{cite}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{algorithmic}
\usepackage{graphicx}
\usepackage{textcomp}
\usepackage{xcolor}

\def\BibTeX{{\rm B\kern-.05em{\sc i\kern-.025em b}\kern-.08em
    T\kern-.1667em\lower.7ex\hbox{E}\kern-.125emX}}
\begin{document}

@@

\begin{abstract}
    % The advancement of technology has been giving contributions to the rapid growth of the use of digital data. In this digital era, lots of physical data have been transformed into the digital ones. One example of the use of digital data is the digital biometric fingerprint data on the Electronic Identity Card (KTPel). Fingerprint matching can take a long time to process if the data is large enough. Thus, there is a need for a parallel fingerprint matching.  Based on this rationale, this paper aims to improve the fingerprint matching performance, in the current state of the art linear solution, by using the Minutia Cylinder-Code (MCC) algorithm in parallel on GPU. Based on the experiment and testing, the proposed solution has a significantly better run time compared to the state of the art linear solution while maintaining the accuracy.

    The advancement of technology has been giving contributions to the rapid growth of the use of digital data. In this digital era, lots of physical data have been transformed into the digital ones. Among those, a lot of confidential data also take place. Thus, security assurance become important as well. One of the entry point of those malicious usages come from server side. Ensuring authority on server utilizing network intrusion detection system (NIDS) could take enormous resource and time. One of the important part of NIDS is string matching part inside the analyzer. Based on this rationale, this paper aims to improve matching speed of detection in order to maximize overall system throughput using GPU instead of CPU. Based on the experiment and testing, the proposed solution has a significantly better run time compared to the state of the art linear solution while maintaining the accuracy. 
\end{abstract}

\begin{IEEEkeywords}
    pattern matching; intrusion detection; GPU; parallel computation; CUDA
\end{IEEEkeywords}

@@

\section{String Matching}
In general, there are two kinds of string matching: single pattern string matching and multi pattern string matching. We will focus into multi pattern string matching. There are several famous algorithms for multi pattern string matching: Aho-Corasick, Commentz-Walter, and Wu-Manber algorithm. This paper will try to implement variation of Aho-Corasick algorithm.

    \subsection{Aho-Corasick Algorithm}
    Aho-Corasick algorithm is kind of string matching algorithm works by matching one string to a dictionary. This dictionary contains all pattern to be matched in form state machine.

    In this algorithm, matching operation is performed by traversing state machine by characters in input string. If any of the final state reached, then input string matched one of the patterns. Matching continued until input string is terminated.

    \begin{figure}[htbp]
        \centerline{\includegraphics[width=0.4\textwidth]{../src/resources/aho-c.png}}
        \caption{Example state machine of Aho-Corasick dictionary.}
        \label{fig}
    \end{figure}

    Aho-Corasick use failure function to perform backtrack after stop at one of the final states. Failure function will redirect this final state to longest pattern prefix which match suffix of the currently matched pattern. Failure function reduce the need of repeating another same sequence which just matched by. By this method, all of containing patterns can be matched by one pass and reducing complexity from $O(mn)$ to $O(n)$.

    In this algorithm, memory lookup become bottleneck. Every traversal operation will fetch one entry from table. Since memory accesses are more expensive than computations, this algorithm is memory bound.

    \subsection{Data Parallel Aho-Corasick}
    In order to extend Aho-Corasick into multithreading, we need to divide the load of matching for each threads. One of the method is by partitioning input into several chunks and then each threads will match this sequence into dictionary. But as we encounter a pattern span multiple chunks at once, it is not recognized on either threads. This problem known as boundary detection problem. So, we need to extend matching by length of the longest pattern. Memory lookup thus increase to $O((n/s + m) * s) = O(n + ms)$ for $s$ is amount of chunks.

    \begin{figure}[htbp]
        \centerline{\includegraphics[width=0.4\textwidth]{../src/resources/boundary.png}}
        \caption{Pattern \textbf{AB} not recognized by both thread 3 and 4}
        \label{fig}
    \end{figure}

    \subsection{Parallel Failureless Aho-Corasick}
    Another method to adapt multithreading Aho-Corasick is to spread threads to all characters byte. Each threads will matching any pattern begin with each coresponding character and terminate if any of final state or no valid transition occurs \cite{lin2013}. The consequence is eavery character will match only one pattern at most.
    Thus, no need to make any failure function. And also boundary detection problem will not occur by using this approach.

    \begin{figure}[htbp]
        \centerline{\includegraphics[width=0.4\textwidth]{../src/resources/pfac.png}}
        \caption{State machine without failure function}
        \label{fig}
    \end{figure}

\section{Related Works}

    \subsection{Utilization of Double Buffering Scheme, Texture Memory and Pinned Memory}

        In this paper, GPU-based matching implementation had been proposed using Aho-Corasick algorithm. Beside utilization of GPU multithreading, there are few adjustment to import Aho-Corasick to GPGPU. State machine built using transition table on 2D table on texture memory. Based on experiment, this implementation improving performance by 19\% \cite{gnort2008}.
        For the memory transfer scheme, double buffering scheme was proposed. Packets will be batched in one buffer. Whenever the buffer gets full, all packets are transferred to the GPU in one operation. Transfer from device to host also done in same way. All result will be batched in the second buffer and get transferred when gets full.
        Another optimization trick used is using pinned memory. Both of the buffer in host will be pinned to reduce swappiness. And pinned memory also enable DMA transfer and reducing computation needs and latency. This design can improve system performance until 3.2 times higher than implementation using multithreading on CPU.

    \subsection{Maximizing GPU Utilization by Pipelining}

        % Sehingga dalam penelitian ini, diajukan beberapa optimasi untuk pencocokan \emph{rule option}. Salah satu optimasi yang diajukan untuk tahap pencocokan yaitu penggunaan \emph{pipelining}. \emph{Pipelining} dilakukan dengan memisahkan \emph{thread} yang digunakan untuk operasi \emph{I/O} dengan operasi analisis. Kemudian pada tiap operasi \emph{I/O}, akan dilakukan pengumpulan paket dalam \emph{batch} sebelum dilakukan transfer antar komponen. Implementasi dilakukan dengan menggunakan basis IDS Snort.

        % Pengujian dilakukan dengan konfigurasi CPU ganda Intel X5680 dengan 12 core dan dua GPU NVIDIA GTX 580. Hasil yang didapatkan menggunakan desain ini mampu mencapai 1,5 sampai 4 kali lipat kinerja Snort. \emph{Latency} dapat menurun hingga 13 mikrodetik pada \emph{offloading} paket awal untuk batch sebesar 1,5 kB. Pada pengujian dengan \emph{traffic} jaringan, rancangan ini mampu mencapai \emph{throughput} 25,2 Gbps pada \emph{input} sebesar 40 Gbps.

        By profiling, it is concluded that bottleneck occured on 3 components: packet acquisition, multi pattern string matching, and option rule matching by regular expression. Moreover, inside string matching component alone,

    \subsection{Failureless Aho-Corasick by Maximizing Shared Memory Usage}

        Salah-satu algoritma yang sering digunakan dalam deteksi pola adalah Aho-Corasick yang bersifat \emph{multi-pattern string matching}. Dalam penelitian yang dilakukan \cite{lin2013}, dilakukan optimasi pada implementasi algoritma Aho-Corasick pada GPU. Implementasi memanfaatkan fitur-fitur pada GPU seperti \emph{memory coalescing}, penghematan transaksi pada \emph{global memory}, dan akses \emph{shared memory} yang bebas \emph{bank-conflict}.

        Desain akan menggunakan pendekatan algoritma AC tanpa \emph{failure function} yaitu PFAC (\emph{Parallel Failureless Aho-Corasick}). Pendekatan ini dapat memaksimalkan penggunaan \emph{thread} pada GPU tanpa menyebabkan \emph{overlapping}. Transaksi ke \emph{global memory} dapat dikurangi dengan memuat transisi \emph{state} kosong ke \emph{shared memory} dan penggunaan \emph{texture memory} yang memanfaatkan \emph{cache}.

        Hasil pengujian dengan CPU Intel Core i7-950 dan GPU NVIDIA GTX 580 menunjukkan peningkatan yang signifikan. Optimasi dengan PFAC pada \emph{multithread} CPU mencapai 7 kali lipat dari AC. Sedangkan PFAC pada GPU dapat mencapai 10 kali lipat daripada AC pada CPU. Masukan didapatkan dari dataset \emph{traffic} kompetisi DEFCON.

\section{Proposed Solution}
    \subsection{Matching Algorithm}
    \subsection{State Machine Implementation}
    \subsection{\emph{Thread} Allocation}
    \subsection{GPU Optimation}

    % \begin{table}[htbp]
    % \caption{Table Type Styles}
    % \begin{center}
    % \begin{tabular}{|c|c|c|c|}
    % \hline
    % \textbf{Table}&\multicolumn{3}{|c|}{\textbf{Table Column Head}} \\
    % \cline{2-4} 
    % \textbf{Head} & \textbf{\textit{Table column subhead}}& \textbf{\textit{Subhead}}& \textbf{\textit{Subhead}} \\
    % \hline
    % copy& More table copy$^{\mathrm{a}}$& &  \\
    % \hline
    % \multicolumn{4}{l}{$^{\mathrm{a}}$Sample of a Table footnote.}
    % \end{tabular}
    % \label{tab1}
    % \end{center}
    % \end{table}

    % \begin{figure}[htbp]
    % \centerline{\includegraphics{fig1.png}}
    % \caption{Example of a figure caption.}
    % \label{fig}
    % \end{figure}

@@

\bibliography{references}
\bibliographystyle{IEEEtran}

### 2018-09-26 src/chapters/forewords.tex
Revisi 3

    \item Ibu Dr. Fazat Nur Azizah S.T, M.Sc. dan Tricya Esterina Widagdo, S.T., M.Sc. selaku para pengajar matakuliah Tugas Akhir yang telah memotivasi saya dan memberikan arahan dalam penyelesaian dan pengerjaan tugas akhir ini.

@@

    \item Khalil Ambiya, Feryandi Nurdiantoro, Nitho Alif Ibadurrahman, dan Muhamad Visat Sutarno yang sempaat menjadi teman seperjuangan dalam mengerjakan tugas akhir ini.

>>>>

    \item Ibu Dr. Fazat Nur Azizah S.T, M.Sc. dan Tricya Esterina Widagdo, S.T., M.Sc. selaku para pengajar matakuliah Tugas Akhir yang telah memotivasi penulis dan memberikan arahan dalam penyelesaian dan pengerjaan tugas akhir ini.

@@

    \item Khalil Ambiya, Feryandi Nurdiantoro, Nitho Alif Ibadurrahman, dan Muhamad Visat Sutarno yang sempat menjadi teman seperjuangan dalam mengerjakan tugas akhir ini.
